## 0. Importing Required Packages
```{r setup, include=FALSE}
library(tidyverse)
library(DESeq2)
library(patchwork)
library(clusterProfiler)
library(enrichplot)
library(pathview)
# Mouse annotation data download
organism <- "org.Mm.eg.db"
# BiocManager::install(organism, character.only = TRUE, force = TRUE)
library(organism, character.only = TRUE)
```

## 1. Loading Samples and Gene Info
For DEG Analysis, we need the following:
* Gene Count Matrix
* Gene Info (For additional/meta gene information)
The count matrix should have the dimensions:
* Cols: Samples
* Rows: Gene ENSEMBL IDs
```{r setup, include=FALSE}
# Loading htseq files
htseq_files <- dir(path = "./htseq", pattern = ".out", full.names = TRUE)
htseq_table_list <- lapply(
  htseq_files,
  function(x) read.table(x, sep = "\t")
)
unique_genes <- unique(unlist(
  # "[[" searches through elements, the 3rd argument is the column number
  lapply(htseq_table_list, "[[", 1),
  use.names = FALSE
))
sample_names <- str_extract(htseq_files, pattern = "SRR[0-9]+")
# Loading gene info
gene_info <- read.table("geneInfo_copy.tsv", sep = "\t")

# creating count matrix
# create empty count matrix
# rownames will be gene names not the gene ID
count_matrix <- structure(
  matrix(
    as.integer(NA),
    nrow = length(unique_genes),
    ncol = length(htseq_files)
  ),
  dimnames = list(unique_genes, sample_names)
)
# fill in the count matrix column-wise
for (i in seq(length(htseq_table_list))) {
  count_matrix[match(htseq_table_list[[i]]$V1, rownames(count_matrix)), i] <-
    htseq_table_list[[i]]$V2
}
# drop non-unique alignments and non-gene entries
drop_classes <- c(
  "__alignment_not_unique", "__no_feature", "__ambiguous",
  "__too_low_aQual", "__not_aligned"
)
if (any(drop_classes %in% rownames(count_matrix))) {
  count_matrix <- count_matrix[-match(drop_classes, rownames(count_matrix)), ]
}
# changing row and column names
# row names will be the gene names instead of gene IDs
# Column names will be sample replication names instead of ENA IDs
colnames(count_matrix) <- c(
  "Naive_Control_Rep_1",
  "Naive_Control_Rep_2",
  "Naive_Control_Rep_3",
  "Naive_DEX_Rep_1",
  "Naive_DEX_Rep_2",
  "Naive_DEX_Rep_3"
)
head(count_matrix)
```

## DESeq Preparation
```{r status, input = FALSE}
# DESeq
col_data <- as.data.frame(c(
  rep("Naive_Control", 3),
  rep("Naive_DEX", 3)
))
rownames(col_data) <- colnames(count_matrix)
colnames(col_data) <- "condition"
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = col_data,
  design = ~condition
)
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]
dds <- DESeq(dds)
normcount <- counts(dds, normalized = TRUE)
res <- results(dds, contrast = c("condition", "Naive_DEX", "Naive_Control"))
```

## 8. GSEA Cluster Profiler
DEG Analasis shows low amount of DEGs (38).
Full ORA may not be useful here
In this case, we will use GSEA cluseter profiler
packages needed to be installed:
- ClusterProfiler
- EnrichPlot
- PathView

# 8.1 Preparation
```{r status, input = FALSE}
# Input preparation
gene_list <- res$log2FoldChange
names(gene_list) <- str_extract(
  rownames(res),
  pattern = "(^[A-Za-z]+[0-9]+).*",
  group = 1
)
gene_list <- na.omit(gene_list)
gene_list <- sort(gene_list, decreasing = TRUE)
head(gene_list)

# gene ontology list
gse <- gseGO(
  geneList = gene_list,
  ont = "ALL",
  keyType = "ENSEMBL",
  nPerm = 10000,
  minGSSize = 3,
  maxGSSize = 800,
  pvalueCutoff = 0.05,
  verbose = TRUE,
  OrgDb = organism,
  pAdjustMethod = "none"
)
```

# 8.2 Dotplot
```{r status, input = FALSE}
require(DOSE)
png(
  "./GSEA_plots/GSEA_1_Enrichment_Dot_Plot.png",
  width = 900,
  height = 900
)
dotplot(gse, showCategory = 10, split = ".sign") + facet_grid(. ~ .sign)
dev.off()
```

# 8.3 Enrichment Map
```{r status, input = FALSE}
png(
  "./GSEA_plots/GSEA_2_Enrichment_Map.png",
  width = 800,
  height = 800,
)
x <- pairwise_termsim(gse, showCategory = 10)
emapplot(
  x,
  showCategory = 10
)
dev.off()
```

# 8.4 Category Network Map
```{r status, input = FALSE}
png("./GSEA_plots/GSEA_3_Enrichment_Cnet_Plot.png")
cnetplot(gse, categorySize = "pvalue", foldChange = gene_list, showCategory = 3)
dev.off()
```

# 8.5 RidgePlot
```{r status, input = FALSE}
png(
  "./GSEA_plots/GSEA_4_Ridge_Plot.png",
  width = 600,
  height = 900
)
ridgeplot(gse) + labs(x = "enrichment distribution")
dev.off()
```

# 8.6 GSEA Analysis
```{r status, input = FALSE}
png(
  "./GSEA_plots/GSEA_5_GSEA_Plot.png"
)
gseaplot(
  gse,
  by = "all",
  geneSetID = 1,
  title = gse$Description[1]
)
dev.off()
```

## 9. KEGG Gene Set Enrichment Analysis
we need to used gseKEGG() function
and convert ID types
```{r status, input = FALSE}
# Convert Gend IDs for gseKEGG function
# Will be losing some genes during this process
ids <- bitr(
  names(gene_list),
  fromType = "ENSEMBL",
  toType = "ENTREZID",
  OrgDb = organism
)
# remove duplicates
dedup_ids <- ids[!duplicated(ids[c("ENSEMBL")]), ]
# Create dataframe for KEGG
kegg_df <- res[str_extract(rownames(res), pattern = "(^[A-Z]+[0-9]+).*", group = 1) %in% dedup_ids$ENSEMBL, ]
kegg_df$Y <- dedup_ids$ENTREZID
kegg_gene_list <- kegg_df$log2FoldChange
names(kegg_gene_list) <- kegg_df$Y
kegg_gene_list <- na.na.omit(kegg_gene_list)
kegg_gene_list <- sort(kegg_gene_list, decreasing = TRUE)

kegg_organism <- "mmu"
kk2 <- gseKEGG(
  geneList = kegg_gene_list,
  organism = kegg_organism,
  nPerm = 10000,
  minGSSize = 3,
  maxGSSize = 800,
  pvalueCutoff = 0.05,
  pAdjustMethod = "none",
  keyType = "ncbi-geneid"
)
```

# 9.1 Kegg Dotplot
```{r status, input = FALSE}
png("./GSEA_plots/KEGG_1_Enrichment_Dot_Plot.png", width = 900, height = 900)
dotplot(
  kk2,
  showCategory = 10,
  title = "Enriched Pathways",
  split = ".sign"
) +
  facet_wrap(. ~ .sign)
dev.off()
```

# 9.2 Enrichment Map
```{r status, input = FALSE}
png(
  "./GSEA_plots/KEGG_2_Enrichment_Map.png",
  width = 800,
  height = 800,
)
x <- pairwise_termsim(kk2, showCategory = 10)
emapplot(
  x,
  showCategory = 10
)
dev.off()
```

# 9.3 Cnet map
```{r status, input = FALSE}
png("./GSEA_plots/KEGG_3_Enrichment_Cnet_Plot.png")
cnetplot(kk2, categorySize = "pvalue", foldChange = gene_list, showCategory = 3)
dev.off()
```

# 8.5 RidgePlot
```{r status, input = FALSE}
png(
  "./GSEA_plots/KEGG_4_Ridge_Plot.png",
  width = 600,
  height = 900
)
ridgeplot(kk2) + labs(x = "enrichment distribution")
dev.off()
```

```{r status, input = FALSE}
png(
  "./GSEA_plots/KEGG_5_GSEA_Plot.png",
  width = 800,
  height = 600
)
gseaplot(
  kk2,
  by = "all",
  geneSetID = 1,
  title = kk2$Description[1]
)
dev.off()
```

# 8.6 KEGG Pathway Analysis
Pathway ID can be found within the dataframe generated by gseKEGG()
```{r status, input = FALSE}
# plot top 3 pathways
top_pathways <- kk2$ID[1:3]
for (path_id in top_pathways) {
  # KEGG Plot
  dme <- pathview(gene.data = kegg_gene_list, pathway.id = path_id, species = kegg_organism)
  # GraphViz Plot
  dme <- pathview(gene.data = kegg_gene_list, pathway.id = path_id, species = kegg_organism, kegg.native = F)
}
```

## EnrichGO Analysis

# preparation
```{r status input = FALSE}
# we want the log2 fold change
original_gene_list <- res$log2FoldChange
# name the vector
names(original_gene_list) <- str_extract(rownames(res), pattern = "(^[A-Z]+[0-9]+).*", group = 1)
# omit any NA values
gene_list <- na.omit(original_gene_list)
# sort the list in decreasing order (required for clusterProfiler)
gene_list <- sort(gene_list, decreasing = TRUE)
# Exctract significant results (padj < 0.05)
sig_genes_res <- subset(res, padj < 0.05)
# From significant results, we want to filter on log2fold change
genes <- sig_genes_res$log2FoldChange
# Name the vector
names(genes) <- str_extract(rownames(sig_genes_res), pattern = "(^[A-Z]+[0-9]+).*", group = 1)
# omit NA values
genes <- na.omit(genes)
# filter on min log2fold change (log2FoldChange > 2)
genes <- names(genes)[abs(genes) > 0.58]

# Ontology options
# BP
# MF
# CC
# ALL

# for this particular example, no enriched terms were found
go_enrich <- enrichGO(
  gene = genes,
  universe = names(gene_list),
  OrgDb = organism,
  keyType = "ENSEMBL",
  readable = T,
  ont = "BP",
  pvalueCutoff = 0.05
)
```

# 1. enrichplot
```{r status, input = FALSE}
enrichplot::upsetplot(go_enrich)
```